---
title: "OP empirical"
output: 
  pdf_document: 
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
n <- 100000
k <- 2
set.seed(2022)
expit <- function(x) exp(x) / (1 + exp(x))
root <- rprojroot::find_root(rprojroot::has_dir(".git"))
r_dir <- file.path(root, "r")
invisible(lapply(list.files(r_dir, full.names = TRUE), source))
```

We take $n = 10^5$ samples and $k = 2$. The probability vector $P(z)$ is given
by $$P(z) = (0.5, 0.2, 0.2, 0.1).$$

```{r}
pz <- c(0.5, 0.2, 0.2, 0.1)
lambda1 <- lambda2 <- 1 # Z_i -> X coefficients
mu1 <- mu2 <- 1 # Z_i -> Y coefficients
beta <- 0.1 # X -> Y coefficient
fi <- 0.3 # fidelity parameter
```

Furthermore, we sample from the SCM as follows:

```{r}
Z <- sample(seq_len(2^k)-1, n, prob = pz, replace = TRUE)
Z <- lapply(Z, function(z) as.integer(intToBits(z)[1:k]))
Z <- do.call(rbind, Z)
X <- rbinom(n, 1, expit(Z %*% c(lambda1, lambda2) - 2))
Y <- rbinom(n, 1, expit(Z %*% c(mu1, mu2) + X * beta - 2))
R <- Z
for (i in seq_len(k)) R[, i] <- Z[, i] * rbinom(n, 1, prob = 1 - fi)
```

```{r, include=FALSE}
# prepare the data frame
dat <- data.frame(X, R, Y)
dat$X <- ifelse(dat$X == 1, 30, 20)
dat$Y <- ifelse(dat$Y == 1, TRUE, FALSE)
names(dat) <- c("bmi", "CHF", "Pulmonary", "death")
dat <- as.data.table(dat)
cmb <- c("CHF", "Pulmonary")
```

Next, we estimate the average treatment effect (ATE) we are trying to recover:
```{r}
ate <- mean(expit(Z %*% c(mu1, mu2) + beta - 2) - expit(Z %*% c(mu1, mu2) - 2))
ate
```

Then, we run the sensitivity analysis using the function `sensitivity_ATE()`
found in the `r/backward-solution.R` script:

```{r}
fixy <- list(list(0, 0), list(0, 0))
fixy[[1]][[1]] <- fixy[[1]][[2]] <- fixy[[2]][[1]] <- fixy[[2]][[2]] <- fi
res <- sensitivity_ATE(fixy, dat, cmb)
cat("P(z) =", res$pz, ",", "ATE =", res$ATE) 
```

Notice that we recovered the ATE and $P(z)$ perfectly, regardless of not having
access to the $Z$ values!

## Over a range of $\phi$ values

We also perform an analysis over a range of $\phi$ values:

```{r, echo=FALSE}

res <- list()
p_range <- seq(0.0, fi, 0.05)
df <- expand.grid(fi = p_range)
for (j in seq_len(nrow(df))) {
  
  fixy <- list(list(0, 0), list(0, 0))
  fixy[[1]][[1]] <- fixy[[1]][[2]] <- fixy[[2]][[1]] <- fixy[[2]][[2]] <- df$fi[j]
  
  res[[j]] <- sensitivity_ATE(fixy, dat, cmb)
  df$adjusted_ATE[j] <- res[[j]][["ATE"]]
  
}

ggplot(df, aes(x = fi, y = adjusted_ATE)) +
  geom_line() + theme_bw() + geom_point() +
  xlab(latex2exp::TeX("Fidelity parameter \\phi")) +
  ylab("Adjusted ATE") +
  ggtitle(latex2exp::TeX("Basic case: $\\phi = \\phi_{x} = \\phi_{xy}$")) +
  scale_y_continuous(labels = scales::percent) +
  geom_hline(yintercept = ate, color = "red") +
  geom_vline(xintercept = 0.3, color = "blue") +
  geom_text(aes(x = 0.05, y = ate+0.001, label = "True ATE"),
            color = "red", angle = 0) +
  geom_text(aes(x = 0.29, y = 0.04, label = "True fidelity"), 
            color = "blue", angle = 90)
```
